---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  echo = FALSE,
  comment = "#>",
  fig.path = "README-"
)
```

# bioset

`bioset` is intended to save you from some dull tasks when dealing with raw data obtained e.g. from a measuring device.

## Installation

You can install `bioset` from github with:

```{r gh-installation, echo = TRUE, eval = FALSE}
# install.packages("devtools")
devtools::install_github("randomchars42/bioset")
```

<!--`bioset` is available on CRAN (<https://CRAN.R-project.org/package=bioset>)-->

## Why? What bioset can do for you

`bioset` lets you:

* import raw data organised in matrices, e.g. measured values of a 9 x 12 bio-assay plate
* calculate concentrations using samples with known concentrations (calibrators) in your dataset
* calculate means and variability for duplicates / triplicates / ...
* convert your concentrations to (more or less) arbitrary units of concentration


```{r, child = "vignettes/introduction.Rmd"}
```

